High-accuracy neutron diffusion calculations based on integral transport theory

### 2019-12-09 docs/EPJP2019/old_version.tex
partial advancement of revision by Tomatis

Since these accurate estimates of the currents are based on a known flux distribution, it was also suggested to execute new diffusion calculations, thus updating iteratively the diffusion coefficients in the global calculation. 
\begin{equation}\label{eq:RM-it}
D^{(k+1)}\rE = - \jtr^{(k)}\rE \left[\bnabla\phi^{(k)}\rE\right]^{-1} \,\, ,
\end{equation}
where $k$ is the iteration index. The use of a tensor notation is needed for the diffusion coefficient in general multidimensional problems.

@@

There are several ways to drive the integral expression for the flux in slab geometry. Most of them start with assuming homogeneous and isotropic scattering and sources, yielding the \emph{Peierls equation} \cite{Davison-1957,Case-1967,Bell-1970,Pomraning-1973,Duderstadt-1979,Lewis-1984}. The derivation of expression for slab geometry (of thickness $a$) is straightforward, yielding
\begin{eqnarray}\label{eq:flux-slab-l-0-no-bndry-terms}
\phi(x,E)&=&\frac{1}{2}\int_0^a { dx'
	E_1\left[\sigma(E)\lvert x-x'\rvert\right] q(x',E) 
}\nonumber \\
&=& \frac{1}{2}\int_0^a { \!\!\! dx'
	E_1\left[\sigma(E)\lvert x-x'\rvert\right]
	\left[
	\int_0^\infty { \!\!\! dE'
		\sigma_s(E\leftarrow E')\phi(x',E') 
	} + S(x',E)
	\right] 
} \,\, ,
\end{eqnarray}
where $E_1(x)$ is a first order exponential integral~\cite{Abramowitz-1964}. Note that this expression does not include the contribution of uncollided neutrons originate from the flux at the boundaries.

Another starting point for the derivation of the integral expression for the flux in slab geometry is to directly integrate along a line the transport equation~\cite{Davison-1957,Pomraning-1973,Duderstadt-1979,Lewis-1984}
\begin{eqnarray}\label{eq:full-ITE-w-bndry-terms}
\psi(\rr,\OO) &=& \int_0^R { dR'
	q(\rr-R'\OO,\OO) e^{-\mcL(\rr,\rr-R'\OO)}
} 
%\nonumber \\
%&+& 
+\psi(\rr-R\OO,\OO) e^{-\mcL(\rr,\rr-R\OO)}
\,\, ,
\end{eqnarray}
where $\mcL(\rr,\rr-R'\OO)$ is the \emph{optical length}, defined as
\begin{equation}\label{eq:opl}
\mcL(\rr,\rr-R'\OO) \equiv 
\int_0^{R'} {\sigma(\rr-R''\OO)dR'} \,\, .
\end{equation}
Assuming isotropic scattering and homogeneous medium in one-dimension for~\eq{eq:full-ITE-w-bndry-terms} recovers~\eq{eq:flux-slab-l-0-no-bndry-terms}.

@@

The angular dependence of the source and the uncollided boundary flux terms is expanded in a series of orthogonal Legendre polynomials
\begin{eqnarray}\label{eq:Q-series}
\psi(0,E,\mu) &=& \sum\limits_{l=0}^\infty {
	\frac{2l+1}{2} \psi_l(0,E) \pl{\mu} 	
} \nonumber \\
\psi(a,E,\mu) &=& \sum\limits_{l=0}^\infty {
	\frac{2l+1}{2} \psi_l(a,E) \pl{\mu} 	
} \,\, .
\end{eqnarray}
The source cross sections are written explicitly as~\cite{Tomatis-2011}
\begin{equation}\label{eq:src-mmnts}
\sigma_l(x,E\leftarrow E')dE' = 
\sigma_{s,l}(x,E\leftarrow E')dE'
+ \delta_{l0}\frac{\chi(E)}{\keff}\nu\sigma_f(x,E')dE' \,\, 
\end{equation}
and the source term is expanded 
\begin{eqnarray}\label{eq:src}
Q(x,E,\mu)&=& \sum\limits_{l=0}^\infty {
	\frac{2l+1}{2} \int_0^\infty { dE'
		\sigma_{s,l}(x,E\leftarrow E')\psi_l(x,E')\pl{\mu} 			
	}
} 
+ \frac{\chi(E)}{2\keff} \int_0^\infty { dE'
	\nu\sigma_f(x,E') \psi_0(x,E')
}  \nonumber \\
&=&\sum\limits_{l=0}^\infty {
	\frac{2l+1}{2} \int_0^\infty { dE'
		\sigma_l(x,E\leftarrow E')\psi_l(x,E')\pl{\mu} 			
	}
} \nonumber \\
&=&\sum\limits_{l=0}^\infty {
\frac{2l+1}{2} \int_0^\infty { dE'
	q_l(x,E\leftarrow E')\pl{\mu} 			
	}
} \nonumber \\
&=&\sum\limits_{l=0}^\infty {
	\frac{2l+1}{2} q_l(x,E)\pl{\mu} 			
} 
\,\, ,
\end{eqnarray}
where the source angular moments are defined as
\begin{equation}\label{eq:q_l}
q_l(x,E) \equiv \int_0^\infty dE' q_l(x,E\leftarrow E') \equiv 
\int_0^\infty dE' \sigma_l(x,E\leftarrow E')\psi_l(x,E') dE' \,\, .
\end{equation}

>>>>

Since these accurate estimates of the currents are based on a known flux distribution, it was also suggested to execute new diffusion calculations, thus updating iteratively the diffusion coefficients in the global calculation. 
\begin{equation}\label{eq:RM-it}
D^{(k+1)}\rE = -\frac{\jtr^{(k)}\rE}{\bnabla\phi^{(k)}\rE} \,\, ,
\end{equation}
where $k$ is the iteration index.

@@

There are several ways to drive the integral expression for the flux in slab geometry. Most of them start with assuming homogeneous and isotropic scattering and sources, yielding the \emph{Peierls equation} \cite{Davison-1957,Case-1967,Bell-1970,Pomraning-1973,Duderstadt-1979,Lewis-1984}. The derivation of expression for slab geometry (of thickness $a$) is straightforward, yielding
\begin{equation}\label{eq:flux-slab-l-0-no-bndry-terms}
\begin{split}
\phi(x,E)=&\frac{1}{2}\int_0^a { dx'
	E_1\left[\sigma(E)\lvert x-x'\rvert\right] q(x',E)
} \text{, with} \\
q(x',E) =& \int_0^\infty { dE'
		\sigma_s(E \leftarrow E')\phi(x',E') + S(x',E)
}
\end{split}
\end{equation}
% commented to overcome an error of latexdiff
%\begin{eqnarray}\label{eq:flux-slab-l-0-no-bndry-terms}
%\phi(x,E)&=&\frac{1}{2}\int_0^a { dx'
%	E_1\left[\sigma(E)\lvert x-x'\rvert\right] q(x',E) 
%}\nonumber \\
%&=& \frac{1}{2}\int_0^a { \!\!\! dx'
%	E_1\left[\sigma(E)\lvert x-x'\rvert\right]
%	\left[
%	\int_0^\infty { \!\!\! dE'
%		\sigma_s(E\leftarrow E')\phi(x',E') 
%	} + S(x',E)
%	\right] 
%} \,\, ,
%\end{eqnarray}
where $E_1(x)$ is a first order exponential integral~\cite{Abramowitz-1964}. Note that this expression does not include the contribution of uncollided neutrons originate from the flux at the boundaries.

Another starting point for the derivation of the integral expression for the flux in slab geometry is to directly integrate along a line the transport equation~\cite{Davison-1957,Pomraning-1973,Duderstadt-1979,Lewis-1984}
\begin{equation}\label{eq:full-ITE-w-bndry-terms}
\psi(\rr,\OO) = \int_0^R { dR'
	q(\rr-R'\OO,\OO) e^{-\mcL(\rr,\rr-R'\OO)}
} + \psi(\rr-R\OO,\OO) e^{-\mcL(\rr,\rr-R\OO)},
\end{equation}
% commented to avoid an error by latexdiff
%\begin{eqnarray}\label{eq:full-ITE-w-bndry-terms}
%\psi(\rr,\OO) &=& \int_0^R { dR'
%	q(\rr-R'\OO,\OO) e^{-\mcL(\rr,\rr-R'\OO)}
%} 
%%\nonumber \\
%%&+& 
%+\psi(\rr-R\OO,\OO) e^{-\mcL(\rr,\rr-R\OO)}
%\,\, ,
%\end{eqnarray}
where $\mcL(\rr,\rr-R'\OO)$ is the \emph{optical length}, defined as
\begin{equation}\label{eq:opl}
\mcL(\rr,\rr-R'\OO) \equiv 
\int_0^{R'} {\sigma(\rr-R''\OO)dR'} \,\, .
\end{equation}
Assuming isotropic scattering and homogeneous medium in one-dimension for~\eq{eq:full-ITE-w-bndry-terms} recovers~\eq{eq:flux-slab-l-0-no-bndry-terms}.

@@

The angular dependence of the source and the uncollided boundary flux terms is expanded in a series of orthogonal Legendre polynomials
\begin{eqnarray}\label{eq:Q-series}
\psi(0,E,\mu) &=& \sum\limits_{l=0}^\infty {
	\frac{2l+1}{2} \psi_l(0,E) \pl{\mu} 	
} \nonumber \\
\psi(a,E,\mu) &=& \sum\limits_{l=0}^\infty {
	\frac{2l+1}{2} \psi_l(a,E) \pl{\mu} 	
} \,\, .
\end{eqnarray}
The source cross sections are written explicitly as~\cite{Tomatis-2011}
\begin{equation}\label{eq:src}
\begin{split}
Q(x,E,\mu)=& \sum\limits_{l=0}^\infty {
	\frac{2l+1}{2} \int_0^\infty { dE'
		\sigma_{s,l}(x,E\leftarrow E')\psi_l(x,E')\pl{\mu} 			
	}
} 
+ \frac{\chi(E)}{2\keff} \int_0^\infty { dE'
	\nu\sigma_f(x,E') \psi_0(x,E')
} \\
=&\sum\limits_{l=0}^\infty {
	\frac{2l+1}{2} \int_0^\infty { dE'
		\sigma_l(x,E\leftarrow E')\psi_l(x,E')\pl{\mu} 			
	}
} \\
=&\sum\limits_{l=0}^\infty {
\frac{2l+1}{2} \int_0^\infty { dE'
	q_l(x,E\leftarrow E')\pl{\mu} 			
	}
} \\
=&\sum\limits_{l=0}^\infty {
	\frac{2l+1}{2} q_l(x,E)\pl{\mu} 			
},
\end{split}
\end{equation}
%\begin{equation}\label{eq:src-mmnts}
%\sigma_l(x,E\leftarrow E')dE' = 
%\sigma_{s,l}(x,E\leftarrow E')dE'
%+ \delta_{l0}\frac{\chi(E)}{\keff}\nu\sigma_f(x,E')dE' \,\, 
%\end{equation}
%and the source term is expanded 
%\begin{eqnarray}\label{eq:src}
%Q(x,E,\mu)&=& \sum\limits_{l=0}^\infty {
%	\frac{2l+1}{2} \int_0^\infty { dE'
%		\sigma_{s,l}(x,E\leftarrow E')\psi_l(x,E')\pl{\mu} 			
%	}
%} 
%+ \frac{\chi(E)}{2\keff} \int_0^\infty { dE'
%	\nu\sigma_f(x,E') \psi_0(x,E')
%}  \nonumber \\
%&=&\sum\limits_{l=0}^\infty {
%	\frac{2l+1}{2} \int_0^\infty { dE'
%		\sigma_l(x,E\leftarrow E')\psi_l(x,E')\pl{\mu} 			
%	}
%} \nonumber \\
%&=&\sum\limits_{l=0}^\infty {
%\frac{2l+1}{2} \int_0^\infty { dE'
%	q_l(x,E\leftarrow E')\pl{\mu} 			
%	}
%} \nonumber \\
%&=&\sum\limits_{l=0}^\infty {
%	\frac{2l+1}{2} q_l(x,E)\pl{\mu} 			
%} 
%\,\, ,
%\end{eqnarray}
where the source angular moments are defined as
\begin{equation}\label{eq:q_l}
q_l(x,E) \equiv \int_0^\infty dE' q_l(x,E\leftarrow E') \equiv 
\int_0^\infty dE' \sigma_l(x,E\leftarrow E')\psi_l(x,E') dE' \,\, .
\end{equation}
